Competing Constructions for German Measure Phrases

### 2017-03-28 Latex/article.tex
massive improvements in stats

\newcommand{\wrt}{w.\,r.\,t.\ }
\newcommand{\Wrt}{W.\,r.\,t.\ }
\newcommand{\eg}{e.\,g.,}
\newcommand{\Eg}{E.\,g.,}
\newcommand{\ie}{i.\,e.,}
\newcommand{\Ie}{I.\,e.,}
\newcommand{\Sub}[1]{\ensuremath{\mathrm{_{#1}}}}
\newcommand{\Sup}[1]{\ensuremath{\mathrm{^{#1}}}}
\newcommand{\Subsf}[1]{\ensuremath{\mathsf{_{#1}}}}
\newcommand{\Supsf}[1]{\ensuremath{\mathsf{^{#1}}}}

@@

This section briefly reviews existing analyses of the \PGCa-\NACa\ alternation and related issues.
I also develop my own analysis and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:externalvalidation}.
%While the constructions and the alternation are standardly covered in descriptive grammars such as \citet[XYZ]{Eisenberg2013a}, the research literature contains little specific discussion of the situation in German.
I discuss four main factors that might influence the alternation:
the different syntactic structures of the variants \NACa\ and \PGCa\ and their syntactic prototypes \NACb\ and \PGCd,
degrees of grammaticalisation of the measure noun associated with the prototypes,
a semantic prototypicality effect of the degrees of grammaticalisation (plurality),
register.

First, I discuss the analysis of the morpho-syntax of German MNPs.
A major interest in descriptive and generative linguistics is to find a plausible analysis of the hierarchical syntactic structure of the constructions in question.
The first question is which noun constitutes the head of the whole construction.
This question was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
However, the MNP-internal structure has more interesting cues to offer if the strong\slash weak inflection patterns of adjectives already mentioned in Section~\ref{sec:descriptive} are taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretistic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners, which is called the \textit{strong} pattern (see Table~\ref{tab:syncretisms}).
Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
They are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected like determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible probabilistic interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase more as a determiner, they \textit{have} to use the \PGCa\ because, if there is a determiner, the PGC (just like in the case of the \PGCd) is the only option.
When they classify the adjective more as an adjective, the kind NP has no determiner, and they \textit{have} to use the \NACa\ (like in the case of the \NACb).%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the GPC as a DP and in the NAC as an NP.}
In a framework that assumes similarity-based probabilistic classification, such as Prototype Theory, we can regard the \PGCd\ as instantiating the PGC prototype and the \NACb\ as the NAC prototype.
The choice between \PGCa\ and \NACa\ is then expected to be influenced by the degree of similarity to either of these prototypes.

Any attempt to substantiate or test this hypothesis is rendered difficult by the fact that the semantic differences between the variants are at best subtle, which means that the prototypes are not associated with clearly detectable differences in meaning.
However, if there is a semantic difference, kind and measure noun lemmas (and maybe semantically of functionally defined classes thereof) could be attracted more strongly to one of the prototypes.
As a result, kind noun lemmas and measure noun lemmas probably appear with different relative frequencies in the prototypical \NACb\ and the prototypical \PGCd.
Their relative frequencies in these non-alternating configurations should then be indicative of their preference to appear in one of the two variants.
This predicted effect will be quantified in Section~\ref{sec:corpusstudies}.

Turning to grammaticalisation now, it is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages per se).
%
%\footnote{Notice that the description of German in \citet[547--549]{Koptjevskaja2001} is vague and partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
%Her example (29b) from \citet[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun (hence, PGC).
%
%\begin{exe}
%  \ex{\label{ex:koptjewskaja}\gll trotz einer Flasche gut-en Wein-s\\
%  in.spite.of one-\textsc{fem.gen} bottle good-\textsc{masc.sg.gen} wine-\textsc{gen}\\
%  \glt in spite of one bottle of good wine}
%\end{exe}
%
%It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \pageref{page:femininesyncretism} and an ambiguous case requirement of the preposition \textit{troz}, hence an instance of NAC.
%Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
%}
%
The alternating constructions in German are both clearly pseudo-partitives, but the grammaticalisation paths uncovered by \cite[esp.\ 526--530]{Koptjevskaja2001} -- which I interpret freely here -- are clearly relevant for an analysis of the alternation.
The grammaticalisation path can start out (in some languages) with constructions involving two referential nouns (not necessarily forming a single NP) and a \textit{separative} meaning as in \textit{(cut) two slices from the cake} \citep[535]{Koptjevskaja2001}.
The \textit{part-of} meaning of true partitives as in \textit{a slice of the cake} represents the first stage of a development wherein the measure noun already tends to loose some semantic content.
The pseudo-partitive stage finally instantiates a \textit{quantity-of} relation, potentially even leading to fully grammaticalised quantifiers such as \textit{a lot}.
In German, the PGC is clearly the older variant \citep{Zimmer2015} with an overt \textit{quantity-of} marker in the form of the genitive suffix.
Also, it still has the potential to form a true partitive (if the kind noun is definite).
Conversely, the NAC completely lacks this ability to form true partitives.
Hence, we can expect more strongly grammaticalised measure nouns to prefer the NAC.
For example, strongly grammaticalised non-referential nouns like \textit{Gramm} `gram' and \textit{Meter} `metre' should occur proportionally more often in the \NACa\ than in the \PGCa.

The question arises whether there might other cues to the degree of grammaticalisation of individual measure nouns or classes of them.
In case studies like \cite{Brems2003,DeclerckBrems2016} syntactic evidence for grammaticalisation plays a prominent role.
In English -- \cite{Brems2003} argues based on a view from \cite{Langacker1991} -- strongly lexicalised and thus semantically bleached measure nouns turn from nouns into quantifiers, leading to a syntactic reanalysis of the NP, wherein the kind noun becomes the head.
The argument is supported by convincing syntactic evidence, for example subject-verb agreement.
As argued above, in German MNPs the measure noun always remains the syntactic head, however, and there is no similar clear criterion to detect grammaticalisation.
Therefore, in the analysis of the corpus finding in Section~\ref{sec:corpusstudies}, effects of different degrees of grammaticalisation will be taken into account by interpreting lemma-specific and lemma class-specific effects.

Furthermore, the grammaticalisation path as described above leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
The best evidence for the relevance of this factor in an ongoing grammaticalisation process is that plural kind nouns almost always (but not exclusively) occur in the \PGCa\ and virtually never in the \NACa\ (see Section~\ref{sec:descriptive}).
More indirectly, however, plurality of the measure noun could also favour the \PGCa\ over the \NACa, and exact quantification in the form of cardinals even more so.%
\footnote{This preference was observed empirically by \cite{Hentschel1993}.}
I provide a highly illustrative example using a measure noun which has virtually developed two variants, one of which grammaticalised to a much higher degree, namely \textit{Haufen} `heap\slash lot'.
The phrase \textit{ein Haufen französische Watte} `a heap\slash lot of French cotton' (\NACa) has essentially two uses.
It can be used to denote an unspecific amount of cotton in the sense of \textit{a lot of French cotton} in situations when the exact quantity of cotton is not of interest.
Under the appropriate circumstances, it could also be used to denote a highly specific quantity of cotton, for example when a cotton merchant always sells her French cotton in \textit{heaps} weighing two stone each.
The former use, however, is highly restricted for \textit{zwei Haufen französischer Watte} `two heaps of French cotton' (\PGCa) -- not `two lots of French cotton'.
Such an NP is less likely to be used in situations where the exact quantity of cotton is irrelevant, even if there are actually two heaps of cotton.
In other words, a precisely quantified measure noun leads to a higher individuation (the creation of discrete portions) of the substance denoted by the kind noun, which is more appropriate for the PGC prototype.
The statistical models in Section~\ref{sec:corpusstudies} will be designed to detect such a subtle effect, if it exists.
%\footnote{In the extreme, some measure nouns cannot be pluralised, such as \textit{eine Menge} `a lot of'.
%Also note that the presentation of the examples in NAC and PGC is suggestive.
%Native speakers will confirm that both NPs sound highly acceptable in both GPC and NAC.
%However, it was made clear from the beginning and throughout that the preferences are indeed subtle.}

Finally, it has been iterated very often that the PGC is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}, where she also refers to descriptive grammars).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial and vernacular variants of German.
Under an integral view, such preferences can be to be part of the construction prototypes.
In the corpus study, the alternation tendencies controlled by this complex factor related to register, medium, and lectal variation will be tested using certain very simple proxies.

@@

This resulting list of mass nouns was used in step (\ref{enum:data:step2}) to bootstrap a list of measure nouns co-occurring with the mass nouns from DECOW14A01.
In order to generate this list, I utilized the fact that a direct sequence of two nouns almost always instantiates the bare-noun NAC if the second noun is a mass noun.
Hence, I searched for all sequences N\Sub{1}N\Sub{2} where N\Sub{2} was one of the mass noun lemmas extracted in step (\ref{enum:data:step1}).
Then, the resulting 100 lists of noun-noun combinations were each sorted by frequency in descending order and sieved manually to remove erroneous hits.
From each of the 100 lists, I also removed noun-noun combinations that had a frequency below 2, except if the individual list would have otherwise been shorter than 20 noun-noun combinations
The result was a list of the most frequent 2,365 individual combinations of a measure noun and a mass noun.

In step (\ref{enum:data:step3}), each of these 2,365 noun–noun combinations was queried in the target constructions (\PGCa\ and \NACa) individually in each of the first ten slices of DECOW (roughly 10 billion tokens).
%The scripts generating the actual queries were crafted so as to make sure that adjectives and nouns were searched in the appropriate forms.
%The CQL query in (\ref{ex:data:query01}) shows an example query for a neuter mass noun like \textit{Wasser} `water' which finds relevant sequences such as the ones shown in (\ref{ex:data:query02}).%
%\footnote{The COW creators used to make corpora available in the IMS Open Corpus Query Workbench \citep{EvertHardie2011}, which has a query syntax informally referred to as CQL.
%Documentation can be obtained at \url{http://cwb.sourceforge.net}.}
%The versions with the genitive singular \textit{Liters} and the dative plural \textit{Litern} are not shown in (2) for space reasons but are also found by the query in (1).
%In terms of context, single sentences containing the target were exported.
%
%\begin{exe}
%  \ex{\label{ex:data:query01} \texttt{\small[ lemma="Liter" ] [ pos="ADJA" \& word=".+(en|em|es)" ] [ lemma="Wasser" ]}}
%  \ex\label{ex:data:query02}
%  \begin{xlist}
%    \ex{\label{ex:data:query02a} Liter [klaren Wassers]\Sub{Gen}}
%    \ex{\label{ex:data:query02b} Liter [klares Wasser]\Sub{Nom\slash Acc}}
%    \ex{\label{ex:data:query02c} Liter [klarem Wasser]\Sub{Dat}}
%  \end{xlist}
%\end{exe}
In order to reduce the sample size for the manual annotation process, the final concordance was sampled from the results of these 2,365 queries.
Since the mass nouns in the sample were distributed according to the usual power law, I used all hits for mass nouns occurring less than one hundred times, but randomly sampled one hundred hits for each mass noun that occurred one hundred or more times.
The final sample contained 6,843 sentences.
Given the careful bootstrapping and sampling procedure described in this section, we can be highly sure that it contains all relevant and reasonably frequent noun–noun combinations in the target constructions.%
\footnote{In a similar fashion, the 100 most frequent measure nouns occurring with plural kind nouns were bootstrapped and queried, resulting in a sample of 1,876 sentences.
%Bootstrapping measure nouns first was more efficient in the plural for reasons to complicated to discuss here, especially because the data was not properly analysed anyway.
As stated in Section~\ref{sec:germanmeasurenps}, the NAC is virtually never used with plural kind nouns, and this sample was not used for full analysis.
%Consequently, it does not make sense to model the influencing factors in the same way as for singular kind nouns.
}

@@

%A mixture of automatic pre-annotation based on the annotations available in DECOW and manual annotations was used.
%Importantly, the case of the measure noun was annotated completely by hand because case syncretisms make automatic case annotations unreliable. 
The full set of manually annotated variables for the main sample is given in Table~\ref{tab:variables}, and I briefly discuss them now.
Notice that \textit{Genitive} is the boolean (dichotomous) response variable (`dependent variable') which is 1 when the PGC is chosen in a given sentence, and 0 when the NAC is chosen.

\begin{table}
  \centering
  \begin{tabular}{llll}
    Unit of reference & Variable                  & Type    & Levels (for factors only)   \\
    \midrule
    Document       & Badness                      & numeric &                             \\
                   & Genitives                    & numeric &                             \\
    Sentence       & \textbf{Genitive (response)} & boolean &                             \\
                   & Matchlength                  & numeric &                             \\
                   & Measureabbreviated           & boolean &                             \\
                   & Measurecase                  & factor  & Nom, Acc, Dat               \\
                   & Measurenumber                & factor  & Pl, Sg                      \\
                   & Minus1pos                    & factor  & A, C, D, O, P               \\
    Kind lemma     & Kindattraction               & numeric &                             \\
                   & Kindconsistency              & factor  & Gaseous, Hard, Immaterial,  \\
                   &                              &         & Liquid, Mixed, Powder, Soft \\
                   & Kindedible                   & boolean &                             \\
                   & Kindfreq                     & numeric &                             \\
                   & Kindint                      & boolean &                             \\
    Measure lemma  & Measureattraction            & numeric &                             \\
                   & Measureclass                 & factor  & Amount, Capacity, Currency, \\
                   &                              &         & Layer, Length, Portion,     \\
                   &                              &         & Rest, Vessel, Weight        \\
                   & Measurefreq                  & numeric &                             \\
  \end{tabular}
  \caption{Annotated variables in the main sample}
  \label{tab:variables}
\end{table}

To capture the influence of register or style mentioned in Section~\ref{sec:analyses}, several proxy variables were used.
At the document level, the DECOW corpus has an annotation for \textit{Badness}.
As described in \cite{SchaeferEa2013}, \textit{Badness} measures how well the distribution of highly frequent short words in the document matches a pre-generated language model for German.
Documents with higher Badness usually contain more incoherent language, shorter sentences, etc.
If the PGC actually favours higher registers and styles, a high \textit{Badness} should be correlated with fewer occurrences of the PGC.
Documents in DECOW14 have also been annotated with a variable called \textit{Genitives}.
The higher the values of this variable, the lower the proportion of genitives among all case-bearing forms is.
While more genitives are also indicative of higher registers, the use of this variable as a regressor in the present study might be considered problematic.
Since the PGC contains a genitive itself, the regressor variable \textit{Genitive} and the document-level variable \textit{Genitives} are not fully independent.
However, since the PGCs make up for only a minute fraction of all genitives, I still use \textit{Genitives} as a regressor with the appropriate caveats.
Since word length is also a good indicator of higher registers and styles, the length of the match in characters (for example 17 for \textit{Kilo tolle Schoki} `a kilo of nice choc' and 49 for \textit{ein Kilogramm vorzüglicher Edelbitter-Schokolade} ` a kilogram of excellent fine dark chocolate') was added as another proxy to register.%
\footnote{The new version of the corpus contains over 100 document-level variables in the spirit of \cite{Biber1988}.
Unfortunately, these annotations were not available when my samples were taken, and even cruder proxies to register than Biberian features (like the length of the query match) had to be used.
All in all, all variables used as indicators of register here have to be interpreted with extreme caution.
}
Furthermore, abbreviated measure nouns like \textit{kg} or \textit{m} might be indicative of less elaborate technical styles, and hence the variable \textit{Measureabbreviated} was included, which is 1 if the measure noun is abbreviated.

>>>>

\newcommand{\wrt}{w.\,r.\,t.\ }
\newcommand{\Wrt}{W.\,r.\,t.\ }
\newcommand{\eg}{e.\,g.,}
\newcommand{\Eg}{E.\,g.,}
\newcommand{\ie}{i.\,e.,}
\newcommand{\Ie}{I.\,e.,}
\newcommand{\Sub}[1]{\ensuremath{\mathrm{_{#1}}}}
\newcommand{\Sup}[1]{\ensuremath{\mathrm{^{#1}}}}
\newcommand{\Subsf}[1]{\ensuremath{\mathsf{_{#1}}}}
\newcommand{\Supsf}[1]{\ensuremath{\mathsf{^{#1}}}}
\newcommand{\pPB}{p\Subsf{PB}}

@@

This section briefly reviews existing analyses of the \PGCa-\NACa\ alternation and related issues.
I also develop my own analysis and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:externalvalidation}.
%While the constructions and the alternation are standardly covered in descriptive grammars such as \citet[XYZ]{Eisenberg2013a}, the research literature contains little specific discussion of the situation in German.
I discuss four main factors that might influence the alternation:
first, the different syntactic structures of the variants \NACa\ and \PGCa\ in relation to their syntactic prototypes \NACb\ and \PGCd;
second, degrees of grammaticalisation of the measure noun associated with the two prototypes;
third, a semantic prototypicality effect of the degrees of grammaticalisation (plurality);
fourth, preferences in different registers.

The first aspect is related to analyses of the morpho-syntax of German MNPs.
A major interest in descriptive and generative linguistics is to find a plausible analysis of the hierarchical syntactic structure of the constructions in question.
The first question is which noun constitutes the head of the whole construction.
This question was answered for the case at hand quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
However, the MNP-internal structure has more interesting cues to offer if the strong\slash weak inflection patterns of adjectives (already mentioned in Section~\ref{sec:descriptive}) are taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretistic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners, however.
This is called the \textit{strong} inflectional pattern (see Table~\ref{tab:syncretisms}).
Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
They are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected like determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible probabilistic interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase more as a determiner, they have to use the \PGCa\ because, if there is a determiner, the PGC (just like in the case of the \PGCd) is the only option.
When they classify the adjective more as an adjective, the kind NP has no determiner, and they have to use the \NACa\ (like in the case of the \NACb).%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the GPC as a DP and in the NAC as an NP.}
In a framework that assumes similarity-based probabilistic classification, such as Prototype Theory, we can regard the \PGCd\ as instantiating the PGC prototype and the \NACb\ as the NAC prototype.
The choice between \PGCa\ and \NACa\ is then expected to be influenced by the degree of similarity to either of these prototypes.

Any attempt to substantiate or test this hypothesis is rendered difficult by the fact that the semantic differences between the variants are at best subtle, which means that the prototypes are not associated with clearly detectable differences in meaning.
However, if there is a semantic difference, kind and measure noun lemmas should be attracted more strongly to one of the prototypes.
As a result, kind noun lemmas and measure noun lemmas are expected to appear with different relative frequencies in the prototypical \NACb\ and the prototypical \PGCd.
Their relative frequencies in these non-alternating configurations should then be indicative of their preference to appear in one of the two alternating variants \NACa\ and \PGCa.
This predicted effect will be quantified in Section~\ref{sec:corpusstudies}.

Turning to grammaticalisation now, it is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages per se).
%
%\footnote{Notice that the description of German in \citet[547--549]{Koptjevskaja2001} is vague and partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
%Her example (29b) from \citet[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun (hence, PGC).
%
%\begin{exe}
%  \ex{\label{ex:koptjewskaja}\gll trotz einer Flasche gut-en Wein-s\\
%  in.spite.of one-\textsc{fem.gen} bottle good-\textsc{masc.sg.gen} wine-\textsc{gen}\\
%  \glt in spite of one bottle of good wine}
%\end{exe}
%
%It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \pageref{page:femininesyncretism} and an ambiguous case requirement of the preposition \textit{troz}, hence an instance of NAC.
%Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
%}
%
The alternating constructions in German are both clearly pseudo-partitives, but the grammaticalisation paths uncovered by \cite[esp.\ 526--530]{Koptjevskaja2001} -- which I interpret freely here -- are clearly relevant for an analysis of the alternation.
The grammaticalisation path can start out (in some languages) with constructions involving two referential nouns (not necessarily forming a single and contiguous NP) and a \textit{separative} meaning as in \textit{(cut) two slices from the cake} \citep[535]{Koptjevskaja2001}.
The \textit{part-of} meaning of true partitives as in \textit{a slice of the cake} represents the first stage of a development wherein the measure noun already tends to loose some semantic content.
The pseudo-partitive stage finally instantiates a \textit{quantity-of} relation, potentially even leading to fully grammaticalised quantifiers such as \textit{a lot}.
In German, the PGC is clearly the older variant \citep{Zimmer2015} with an overt \textit{quantity-of} marker in the form of the genitive suffix.
Also, it still has the potential to form a true partitive (if the kind noun is definite).
Conversely, the NAC completely lacks this ability to form true partitives, and it forms a more restrictive environment with less combinatory and semantic flexibility.
Hence, we can expect more strongly grammaticalised measure nouns to prefer the NAC.
For example, strongly grammaticalised non-referential nouns like \textit{Gramm} `gram' and \textit{Meter} `metre' should occur proportionally more often in the \NACa\ than in the \PGCa.%
\footnote{In case studies like \cite{Brems2003,DeclerckBrems2016} syntactic evidence for grammaticalisation plays a prominent role.
In English -- \cite{Brems2003} argues based on a view from \cite{Langacker1991} -- strongly lexicalised and thus semantically bleached measure nouns turn from nouns into quantifiers, leading to a syntactic reanalysis of the NP, wherein the kind noun becomes the head.
The argument is supported by convincing syntactic evidence, for example subject-verb agreement.
As argued above, in German MNPs the measure noun always remains the syntactic head, and there is no similarly handy criterion to diagnose a high degree of grammaticalisation.}

Furthermore, the grammaticalisation path as described above leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
The best evidence for the relevance of this factor in an ongoing grammaticalisation process is that plural kind nouns almost always (but not exclusively) occur in the \PGCa\ and virtually never in the \NACa\ (see Section~\ref{sec:descriptive}).
More indirectly, however, plurality of the measure noun could also favour the \PGCa\ over the \NACa, and exact quantification in the form of cardinals even more so.%
\footnote{Without offering a theoretical analysis, such a preference was observed empirically by \cite{Hentschel1993}.}
There are highly illustrative examples with measure nouns that have virtually developed two variants, one of which grammaticalised to a much higher degree, \eg\ \textit{Haufen} `heap\slash lot'.
The phrase \textit{ein Haufen französische Watte} `a heap\slash lot of French cotton' (\NACa) has two uses.
It can be used to denote an unspecific amount of cotton in the sense of \textit{a lot of French cotton} in situations when the exact quantity of cotton is not of interest.
Under the appropriate circumstances, it could also be used to denote a heap-shaped, potentially highly specific quantity of cotton, for example when a cotton merchant always sells her French cotton in \textit{heaps} weighing two stone each.
The former use, however, is highly restricted for \textit{zwei Haufen französischer Watte} `two heaps of French cotton' (\PGCa) -- not `two lots of French cotton'.
Such an NP is less likely to be used in situations where the exact quantity of cotton is irrelevant, even if there are actually two heaps of cotton.
In other words, a precisely quantified measure noun leads to a higher individuation (the creation of discrete portions) of the substance denoted by the kind noun, which is more appropriate for the PGC prototype.
The statistical models in Section~\ref{sec:corpusstudies} will be designed to detect such a subtle effect, if it exists.
%\footnote{In the extreme, some measure nouns cannot be pluralised, such as \textit{eine Menge} `a lot of'.
%Also note that the presentation of the examples in NAC and PGC is suggestive.
%Native speakers will confirm that both NPs sound highly acceptable in both GPC and NAC.
%However, it was made clear from the beginning and throughout that the preferences are indeed subtle.}

Finally, it has been iterated very often that the PGC is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}, where she also refers to descriptive grammars).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial and vernacular variants of German.
Under an integral view, such preferences can be to be part of the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modelled, even if only using some very simple proxies.

@@

This resulting list of mass nouns was used in step (\ref{enum:data:step2}) to bootstrap a list of measure nouns co-occurring with the mass nouns from DECOW14A01.
In order to generate this list, I utilized the fact that a direct sequence of two nouns almost always instantiates the bare-noun NAC if the second noun is a mass noun.
Hence, I searched for all sequences N\Sub{1}N\Sub{2} where N\Sub{2} was one of the mass noun lemmas extracted in step (\ref{enum:data:step1}).
Then, the resulting 100 lists of noun-noun combinations were each sorted by frequency in descending order and sieved manually to remove erroneous hits.
From each of the 100 lists, I also removed noun-noun combinations that had a frequency below 2, except if the individual list would have otherwise been shorter than 20 noun-noun combinations.
The result was a list of the most frequent 2,365 individual combinations of a measure noun and a mass noun.

In step (\ref{enum:data:step3}), each of these 2,365 noun–noun combinations was queried in the target constructions (\PGCa\ and \NACa) individually in each of the first ten slices of DECOW (roughly 10 billion tokens).
%The scripts generating the actual queries were crafted so as to make sure that adjectives and nouns were searched in the appropriate forms.
%The CQL query in (\ref{ex:data:query01}) shows an example query for a neuter mass noun like \textit{Wasser} `water' which finds relevant sequences such as the ones shown in (\ref{ex:data:query02}).%
%\footnote{The COW creators used to make corpora available in the IMS Open Corpus Query Workbench \citep{EvertHardie2011}, which has a query syntax informally referred to as CQL.
%Documentation can be obtained at \url{http://cwb.sourceforge.net}.}
%The versions with the genitive singular \textit{Liters} and the dative plural \textit{Litern} are not shown in (2) for space reasons but are also found by the query in (1).
%In terms of context, single sentences containing the target were exported.
%
%\begin{exe}
%  \ex{\label{ex:data:query01} \texttt{\small[ lemma="Liter" ] [ pos="ADJA" \& word=".+(en|em|es)" ] [ lemma="Wasser" ]}}
%  \ex\label{ex:data:query02}
%  \begin{xlist}
%    \ex{\label{ex:data:query02a} Liter [klaren Wassers]\Sub{Gen}}
%    \ex{\label{ex:data:query02b} Liter [klares Wasser]\Sub{Nom\slash Acc}}
%    \ex{\label{ex:data:query02c} Liter [klarem Wasser]\Sub{Dat}}
%  \end{xlist}
%\end{exe}
In order to reduce the sample size for the manual annotation process, the final concordance was sampled from the results of these 2,365 queries.
Since the mass nouns in the sample were distributed according to the usual power law, I used all hits for mass nouns occurring less than one hundred times, but randomly sampled one hundred hits for each mass noun that occurred one hundred or more times.
The final sample contained 6,843 sentences, which was reduced to 5,063 in the manual annotation process due to removal of noisy material, erroneous hits and uninformative cases where the measure noun was in the genitive, in which case the \NACa\ cannot be distinguished from the \PGCa.
Given the careful bootstrapping and sampling procedure described in this section, we can be highly sure that it contains all relevant and reasonably frequent noun–noun combinations in the target constructions.%
\footnote{In a similar fashion, the 100 most frequent measure nouns occurring with plural kind nouns were bootstrapped and queried, resulting in a sample of 1,876 sentences.
%Bootstrapping measure nouns first was more efficient in the plural for reasons to complicated to discuss here, especially because the data was not properly analysed anyway.
As stated in Section~\ref{sec:germanmeasurenps}, the NAC is virtually never used with plural kind nouns, and this sample was not used for full analysis.
%Consequently, it does not make sense to model the influencing factors in the same way as for singular kind nouns.
}

@@

The full set of manually annotated variables for the main sample is given in Table~\ref{tab:variables}, and I briefly discuss them now.%
\footnote{All numeric variables were also z-transformed (\ie, centered to the mean and rescaled such that they have a standard deviation of $1$) to facilitate their interpretation in the regression models reported in the next section.}
Notice that \textit{Construction} is the response variable (or `dependent variable') with the values \textit{PGCa} and \textit{NACa}.

\begin{table}
  \centering
  \begin{tabular}{llll}
    Unit of reference & Variable                      & Type    & Levels (for factors only)    \\
    \midrule
    Document       & Badness                          & numeric &                              \\
                   & Genitives                        & numeric &                              \\
    Sentence       & \textbf{Construction (response)} & factor  & NACa, PGCa                   \\
                   & Measurecase                      & factor  & Nom, Acc, Dat                \\
                   & Measurenumber                    & factor  & Pl, Sg                       \\
                   & Leftcontext                      & factor  & C, D, O, A, P                \\
    Kind lemma     & Kindattraction                   & numeric &                              \\
                   & Kindfreq                         & numeric &                              \\
    Measure lemma  & Measureattraction                & numeric &                              \\
                   & Measureclass                     & factor  & Physical, Container, Amount, \\
                   &                                  &         & Portion, Rest                \\
                   & Measurefreq                      & numeric &                              \\
  \end{tabular}
  \caption{Annotated variables for the main sample}
  \label{tab:variables}
\end{table}

The variables \textit{Kindattraction} and \textit{Measureattraction} encode the ratio with which a given kind noun lemma or measure noun lemma occurs in the \NACb\ and the \PGCd.
They were calculated from the auxiliary samples described at the end of Section~\ref{sec:gettingdata} as a log-transformed quotient.
The higher the value, the more often the noun occurs in the \PGCa (proportionally).
\textit{Kindfreq} and \textit{Measurefreq} are the logarithm-transformed frequencies per 1,000,000 words of each lemma, extracted from the frequency lists distributed by the DECOW corpus creators on their web page.

In Section~\ref{sec:analyses}, it was hypothesised that classes of measure lemmas might have different preferences for the two variants.
To capture this, class information was annotated for measure lemmas.
The classification was designed inspired by the list in \cite[530]{Koptjevskaja2001}, but due to the frequencies of the respective nouns, very coarse classes were necessary.
With typical examples and the number of data points in the final sample, the classes are:
\textit{Physical} (abstract precisely measurable units such as \textit{Liter} `litre', \textit{Meter} `metre', \textit{Gramm} `gram'; size 1,968),
\textit{Container} (\textit{Eimer} `bucket'; size 740), 
\textit{Amount} (\textit{Menge} `amount'; size 1,364), 
\textit{Portion} (natural portions like \textit{Happen} `bite' or \textit{Krümel} `crumb'; size 713).
The few lemmas that did not fit into either of these classes were labeled \textit{Rest} (size 278).

The variable \textit{Leftcontext} encodes the coarse-grained part-of-speech of the word preceding the measure noun.
The level names stand for (again with class sizes in the final sample) \textit{cardinal} (size 1,939), \textit{determiner} (size 1,971), \textit{other} (size 221), \textit{adjective} (size 845), and \textit{preposition} (size 87).
The purpose of this variable is mainly to test whether cardinals really favour the \PGCa\ as hypothesised in Section~\ref{sec:analyses}.
Along the same lines, \textit{Measurenumber} encodes the grammatical case and number of the measure noun.
